---
title: "odeintr"
author: "Timothy H. Keitt"
date: "Feb 14, 2017"
output:
  html_document:
    keep_md: yes
---

[![CRAN Version](https://www.r-pkg.org/badges/version/odeintr)](https://www.r-pkg.org/badges/version/odeintr) [![CRAN Downloads](https://cranlogs.r-pkg.org/badges/odeintr)](https://cran.r-project.org/package=odeintr) [![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/thk686/odeintr?branch=master&svg=true)](https://ci.appveyor.com/project/thk686/odeintr)
[![Travis-CI Build Status](https://travis-ci.org/thk686/odeintr.svg?branch=master)](https://travis-ci.org/thk686/odeintr) 

The odeintr is package for integrating differential equations in R. The integration engine is
the [Boost odeint package](https://headmyshoulder.github.io/odeint-v2/).

### Features

1. Simple specification of the ODE system
1. Named, dynamic run-time setable system parameters
1. Intelligent defaults, easily overridden, used throughout
1. A wide range of integration methods available for compiled system (see [stepper types](https://www.boost.org/doc/libs/1_58_0/libs/numeric/odeint/doc/html/boost_numeric_odeint/odeint_in_detail/steppers.html#boost_numeric_odeint.odeint_in_detail.steppers.stepper_overview))
1. Fully automated compilation of ODE system specified in C++
1. ~~Simple openmp vectorization of large systems~~ (Broken in latest Boost release)
1. Results returned as a simple data frame ready for analysis and plotting
1. Ability to specify a custom observer in R that can return arbitrary data
1. Three options for calling the observer: at regular intervals, after each update step or at specified times
1. Ability to alter system state and restart simulations where you left off
1. Can compile an implicit solver with symbolic evaluation of the Jacobian
1. You can easily save and edit the generated C++ code

### Installation

```
install.packages(odeintr)                   # released
devtools::install_github("thk686/odeintr")  # development
```

### Examples

```{r, fig.width = 5, fig.height = 5}
library(odeintr)
dxdt = function(x, t) x * (1 - x)
system.time({x = integrate_sys(dxdt, 0.001, 15, 0.01)})
plot(x, type = "l", lwd = 3, col = "steelblue", main = "Logistic Growth")
compile_sys("logistic", "dxdt = x * (1 - x)")
system.time({x = logistic(0.001, 15, 0.01)})
plot(x, type = "l", lwd = "3", col = "steelblue", main = "Logistic Growth")
points(logistic_at(0.001, sort(runif(10, 0, 15)), 0.01), col = "darkred")
```

```{r, fig.width = 5, fig.height = 5}
dxdt = function(x, t) c(x[1] - x[1] * x[2], x[1] * x[2] - x[2])
obs = function(x, t) c(Prey = x[1], Predator = x[2], Ratio = x[1] / x[2])
system.time({x = integrate_sys(dxdt, rep(2, 2), 20, 0.01, observer = obs)})
plot(x[, c(2, 3)], type = "l", lwd = 2, col = "steelblue", main = "Lotka-Volterra Phase Plot")
plot(x[, c(1, 4)], type = "l", lwd = 2, col = "steelblue", main = "Prey-Predator Ratio")
```

```{r, fig.width = 5, fig.height = 5}
# C++ code
Lorenz.sys = '
  dxdt[0] = 10.0 * (x[1] - x[0]);
  dxdt[1] = 28.0 * x[0] - x[1] - x[0] * x[2];
  dxdt[2] = -8.0 / 3.0 * x[2] + x[0] * x[1];
  ' # Lorenz.sys
compile_sys("lorenz", Lorenz.sys)
system.time({x = lorenz(rep(1, 3), 100, 0.001)})
plot(x[, c(2, 4)], type = 'l', col = "steelblue", main = "Lorenz Attractor")
```

```{r, fig.width = 5, fig.height = 5}
VdP.sys = '
dxdt[0] = x[1];
dxdt[1] = 2.0 * (1 - x[0] * x[0]) * x[1] - x[0];
' # Vdp.sys
compile_sys("vanderpol", VdP.sys, method = "bsd") # Bulirsch-Stoer
system.time({x = vanderpol(rep(1e-4, 2), 100, 0.01)})
par(mfrow = c(2, 2), mar = rep(0.5, 4), oma = rep(5, 4), xpd = NA)
make.plot = function(xy, xlab = NA, ylab = NA)
  plot(xy, col = "steelblue", lwd = 2, type = "l",
       axes = FALSE, xlab = xlab, ylab = ylab)
plot.new()
make.plot(x[, c(3, 1)]); axis(3); axis(4)
make.plot(x[, c(1, 2)], "Time", "X1"); axis(1); axis(2)
make.plot(x[, c(3, 2)], "X2"); axis(1); axis(4)
title(main = "Van der Pol Oscillator", outer = TRUE)
```

```{r, fig.width = 5, fig.height = 5}
# Use a dynamic parameter
VdP.sys = '
dxdt[0] = x[1];
dxdt[1] = mu * (1 - x[0] * x[0]) * x[1] - x[0];
' # Vdp.sys
compile_sys("vpol2", VdP.sys, "mu", method = "bsd")
par(mfrow = c(2, 2), mar = rep(1, 4), oma = rep(3, 4), xpd = NA)
for (mu in seq(0.5, 2, len = 4))
{
  vpol2_set_params(mu = mu)
  x = vpol2(rep(1e-4, 2), 100, 0.01)
  make.plot(x[, 2:3]); box()
  title(paste("mu =", round(mu, 2)))
}
title("Van der Pol Oscillator Parameter Sweep", outer = TRUE)
title(xlab = "X1", ylab = "X2", line = 0, outer = TRUE)
```

```{r, fig.width = 5, fig.height = 5}
# Stiff example from odeint docs
Stiff.sys = '
  dxdt[0] = -101.0 * x[0] - 100.0 * x[1];
  dxdt[1] = x[0];
' # Stiff.sys
cat(JacobianCpp(Stiff.sys))
compile_implicit("stiff", Stiff.sys)
x = stiff(c(2, 1), 5, 0.001)
plot(x[, 1:2], type = "l", lwd = 2, col = "steelblue")
lines(x[, c(1, 3)], lwd = 2, col = "darkred")
```

The example below is not working with the latest version of odeint that comes with the BH package. I've submitted and issue on github.

```{r, fig.width = 5, fig.height = 5}
# Robertson chemical kinetics problem
# Robertson = '
# dxdt[0] = -alpha * x[0] + beta * x[1] * x[2];
# dxdt[1] = alpha * x[0] - beta * x[1] * x[2] - gamma * x[1] * x[1];
# dxdt[2] = gamma * x[1] * x[1];
# ' # Robertson
# pars = c(alpha = 0.04, beta = 1e4, gamma = 3e7)
# init.cond = c(1, 0, 0)
# cat(JacobianCpp(Robertson))
# compile_implicit("robertson", Robertson, pars, TRUE)
# at = 10 ^ seq(-5, 5, len = 400)
# x = robertson_at(init.cond, at)
# par(mfrow = c(3, 1), mar = rep(0.5, 4), oma = rep(5, 4), xpd = NA)
# plot(x[, 1:2], type = "l", lwd = 3,
#      col = "steelblue", log = "x", axes = F, xlab = NA)
# axis(2); box()
# plot(x[, c(1, 3)], type = "l", lwd = 3,
#      col = "steelblue", log = "x", axes = F, xlab = NA)
# axis(4); box()
# plot(x[, c(1, 4)], type = "l", lwd = 3,
#      col = "steelblue", log = "x", axes = F)
# axis(2); axis(1); box()
```

It is important to understand that the only thing that ```odeintr``` does is to generate C++ code and compile it using Rcpp. That means you can use the generated code as a starting point for your project and modify as you wish.

```{r}
the_code = compile_sys("logitest", "dxdt = x * (1 - x)", compile = FALSE)
cat(the_code)
```

### Performance

Because ODEINT is a header-only library, the entire integration path is exposed to the compiler. That means your system functions can be inlined with the integration code, loops unrolled, etc. It will help if you enable optimization in your compiler. Use "-O3" with gcc. See the R documentation on the user Makevars file. (Odeintr now provides a convenient function to set the compiler optimization level.)

The Lorenz and Van der Pol examples above show about 10 million observer calls per second.

### To Do

1. ~~Add additional integration methods from odeint~~
1. ~~Extend customized observer to compiled code~~
1. ~~Allow user to set error tolerances in compiled code~~
1. ~~Allow user to set error tolerances for system defined in R~~
1. ~~Expose implicit solver methods~~
1. ~~Compute Jacobian symbolically~~
1. ~~Convenient dynamic parameter settings~~
1. ~~Install emitted function in a new environment~~
1. Add more control of openmp threading
1. Allow custom state type

Pull requests are welcome.

