---
title: "Integrating differential equations in R with odeintr"
author: "Timothy H. Keitt"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrating differential equations in R with odeintr}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Differential equations are the workhorses of the modeling world and find broad application across the sciences, engineering and mathematics. Nearly all modern scientific computational environments provide facilities to compute numerical solutions to systems of differential equations and R is no exception^[The [differential equations task view on CRAN](http://cran.r-project.org/web/views/DifferentialEquations.html) lists many of the available packages]. For example, the deSolve package of Soetaert, Petzoldt and Setzer (cite) wraps a number of legacy Fortran codes for solving differential equations, notably Hindmarsh's lsode and vode solvers. These are well-known, highly-efficient solvers covering a range of common modeling problems.

Differential equation solvers approximate continuous-time solutions by accumulation over discrete iterated computations of the derivatives, and in the case of certain implicit solvers, the Jacobian. When working with large complex model systems, it is desirable to specify these functions in a high-performance compiled language such as Fortran or C++ owing to the computational overhead of the R interpreter. Currently in R, the facilities for compiling the derivative and Jacobian functions are complex and expose the user to aspects of the C-level R internal programming model. The goal of odeintr is to make compilation of differential equation solvers in C++ simple and easy. In odeintr, the user is only required to specify the minimal set of C++ expression to compute the time derivatives, and the Jacobian matrix when required. From these expressions, a custom solver is compiled on-the-fly and installed in the R environment^[Users may also wish to explore the [ccSolve](https://github.com/karlines/ccSolve) and [rodeint](https://github.com/richfitz/rodeint) projects. I became aware of these projects after the initial version of odeintr was completed].

It accomplishes this by using the inline-compilation capabilities of Rcpp

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` setion of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
